---
title: "R Shiny Cookbook"
author: "Markus Lang"
site: bookdown::bookdown_site
documentclass: book
output:
  bookdown::gitbook: default
  #bookdown::pdf_book: default
---

<hr>
# Preface

The primary reason for me to start this project was to document my own work. I develop professional apps and dashboards with R-Shiny. I work in a small team on these apps and know since my studies how to handle statistics and numbers, but no programming principles like "clean code". With more and more experience, it seems absolutely necessary for me to write documentation of my work instead of searching for hours in old projects for code snippets that I once used. 

## What this book will cover

Although R-Shiny allows you to develop apps without HTML, CSS, and JavaScript knowledge, the real power comes from integrating these technologies. Especially the communication between R and JavaScript makes it possible to create really fantastic apps. <br>In addition to the core technologies there are other languages and tools that make my everyday life easier and bring Shiny to life outside of shinyapps.io: Linux (Shell/Bash), Nginx, Git and some other tools. <br>
Therefore, the book does not deal with the basics anymore, neither in R nor in Shiny, but should help people who already have experience with R Shiny. Furthermore, this is not a how-to book, but rather a collection of best practices. In addition to own solutions for problems also links to webinars, articles or Stackoverflow posts are included. If you want to contribute something yourself, be it with code or interesting links just write me a message to langmarkus@hotmail.com.

# Data Wrangling
This chapter does not deal with the data wrangling itself, but with how to calculate the numbers as quickly as possible from a data set that has already been cleaned up, and how to display them in the dashboard. The reason for this is that R-Shiny applications should always load fast readable and clean data sets, because this is an essential factor for a good performance of the app.

## Data Wrangling with dplyr
Dashboards are often used to quickly and easily display differences between groups. This workflow perfectly fits to using dplyr, which with its simple pipe syntax allows you to create code that is easy for others to read. 


```{r, message=FALSE, warning=FALSE}
library(dplyr)
result <- mtcars %>%
  group_by(cyl) %>%
  summarise(meanmpg = mean(mpg, na.rm = TRUE))
print(result)
```
Easy to write, easy to read! But in Shiny Apps, but often might have to write code like this many times, for example when you want to group by different variables. Maybe you have heard this:

> When you’ve written the same code 3 times, write a function
>
> -- <cite>Hadley Wickham</cite>

Nothing easier than this:

```{r, message=FALSE, warning=FALSE, error=TRUE}
grouped_mean <- function(data, group, sumvar) {
    result <- data %>%
        group_by(group) %>%
        summarise(rumresult = mean(sumvar, na.rm = TRUE))
    return(result)
}

grouped_mean(mtcars, cyl, mpg)
```
The error shows, that dplyr was not able to find the grouping variable. The reason for this is something called non-standard-evaluation (NSE). This expression is a term that covers everything that does not follow the standard rules of R evaluation. NSE is archieved with meta programming techniques which allow the programmer to manipulate the way R evaluates functions. NSE allows to write code without referring to the dataframe: `group_by(cyl)` instead of `group_by(mtcars$cyl)`. Get more and in depth information about this topic here:

[Advanced R by Hadley Wickham](https://adv-r.hadley.nz/) <br>
[Metaprogramming in R: Advanced Statistical Programming for Data Science, Analysis and Finance by Thomas Mailund](Wickham]https://www.amazon.de/Metaprogramming-Advanced-Statistical-Programming-Analysis/dp/1484228804/ref=sr_1_1?__mk_de_DE=%C3%85M%C3%85%C5%BD%C3%95%C3%91&keywords=metaprogramming+r&qid=1555258801&s=gateway&sr=8-1)

This ease of use comes with the drawback that dplyr is refencially not transparent. For functions you just could use base R functions, but dplyr also preserves functions which use standard evaluation.

### NSE and Tidy evaluation
Before version 0.8.0 most dplyr functions had a "_" equivalent. Since dplyr 0.8.0 there is a new and easier option to use dplyr code in custom functions. dplyr 0.8.0 allows to use something called tidy evaluation. The key principle behind tidy evaluation, instead of evaluating the arguments, is to quote arguments which refer to data frame columns and unquote them in the process. 

Quoting a variable is archieved with the `enquo()` function. The quoted arguments get passed to the function get passed to the function and will get evaluated when being unquoted with `!!`

```{r, message=FALSE, warning=FALSE}
grouped_mean <- function(data, group, sumvar) {
    
    group <- enquo(group)
    sumvar <- enquo(sumvar)
  
    result <- data %>%
        group_by(!!group) %>%
        summarise(sumresult = mean(!!sumvar, na.rm = TRUE))
    return(result)
}

grouped_mean(mtcars, cyl, mpg)
```

This works perfectly fine and allows you to get rid of repetitive code. The topic tidy evaluation is of course much more extensive than what was shown in this chapter. Further information on this topic can be found here:

[dplyr 0.8.0 documentation](https://dplyr.tidyverse.org/articles/programming.html#quoting) <br>
[Tidy evaluation by Hadley Wickham](https://tidyeval.tidyverse.org/index.html)

# Interactive Data Visualisation

R has long offered one of the best visualization packages for data science - ggplot by Hadley Wickham. As nice and easy as it is to use, the expectation of users on websites is that you can interact with the graphics. Many libraries have been written with JavaScript for this purpose, and many R-wrappers have been written for them. In my personal opinion the best visualization package is HighcharteR. It is a wrapper for Highcharts.js and allows to create very nice visualizations quickly and easily, which still have a lot of possibilities to customize the output. The documentation of the website is really excellent. Unfortunately highcharts does not offer a crossfilter option, but with a workaround JavaScript this function can be emulated to a certain degree. 


## The highcharter Package
The Highcharter Package was written by JbKunst (thanks a lot for that!). The following would like to show different types of charts.

### Barchart

### Scatterplot

### Linechart

# HTML

# CSS

# Bootstrap

# JavaScript

# Git

# Linux Basics
To be able to offer R-Shiny dashboards to customers or use it for your own portfolio website without having the .shinyapps.io extension, it is worthwhile to buy your own domain and a web server (in 2019 a good investment for everyone in tech). Domains are available for 20 dollar per year, a webserver is available from 5 dollar per month. I personally use and recommend digitalocean because the server is very easy to use and a complete virtual machine can be set up within a few minutes.

## Set up R-Shiny on a Linux Ubuntu 18.04 VM
First thing to do on your vm is to set up a non root-user.
```console
adduser markus
gpasswd -a markus sudo

su - markus
```

Simple rule - never use root user to manipulate anything on your server. Always use a "superuser".

Next Step is to install a webserver, where shiny apps can run. The two most popular are apache and nginx. Nginx seems to be better suited and better documented in combination with shiny.
```console
sudo apt-get update
sudo apt-get -y install nginx
```

Nginx works as a service. The most often used commands are start, stop and restart which should be selfexplanatory. Before installing R, you have to add the KEY and Repository

```console
sudo apt-key adv --keyserver keyserver.ubuntu.com --recv-keys E298A3A825C0D65DFD57CBB651716619E084DAB9
sudo add-apt-repository 'deb https://cloud.r-project.org/bin/linux/ubuntu bionic-cran35/'
sudo apt update
```

Now install R
```console
sudo apt install r-base
```
To see if everything worked fine, start R:
```console
sudo R
```
## Installing packages
I recommend to never install packages inside R, but always from the command line. Reason for this is that packages should be available to every user. When you install packages in R, they will be saved in the personal library and others users can not use them. So always install the packages with ```console sudo su```

Now install shiny with the following command:

```console
sudo su - -c "R -e \"install.packages('shiny, repos='http://cran.rstudio.com/')\""
```

Now get the latest version of shiny server:

```console
wget https://download3.rstudio.org/ubuntu-12.04/x86_64/shiny-server-1.5.6.875-amd64.deb
sudo gdebi shiny-server-1.5.6.875-amd64.deb
```


## Use Linux to automatically send reports
Linux can offer great opporunities to run scripts from a scheduler called cron that allows you to execute scripts at repeating points in time. This is great for pulling data from API´s and writing it to a database that will be accessed by an app the next day. Even things like server updates can be moved to times of day where they don't get in touch with active users. Cron is preinstalled and can be accessed with:
```console
crontab -e
```

Cron commands look liket his
```console
0 5 * * 1 tar -zcf /var/backups/home.tgz /home/
```
This websites helps you with the settings how often you want to call a cron job:
https://crontab.guru/every-10-minutes

Some issues which caused me a headache when my scripts were not executed, but worked when i executed them from the command line:
- Permission: You always have set the settings to allow the script being executed. The following command allows the current user to read, write and execute the file.
```console
sudo chmod 777 /path/script.R
```
Another big issue was abouts paths: It seems to be a good solution to always use absolute paths if you want to run cron (point to the folder where your script lives). 

```console 
# Bad
*/30 * * * * script.sh > listener2.log 2>&1
# Good
*/30 * * * * /bin/sh /home/markus/Spielplatz/script.sh > /tmp/listener2.log 2>&
```
The problem with cron is that errors are silent, so check these two issues first if your script does not work as expected. There might be other problems of course, but I do not use cron so often to be familiar with all of then.

## Daily reports with mails (Gmail):

Sometimes it´s more suitable to send reports to your clients, may it be on a daily, weekly or monthly basis. 
To create a report, you need a markdown.Rmd file and a second file (script.R), whichs calls the second markdown file. Here is a basic example how this works, both codes worked fine in my example.

```{r, eval=FALSE, error=FALSE} 
library(knitr)
library(rmarkdown)

knit(input = "/home/user/stockreport.Rmd") 
render(input = "/home/user/stockreport.Rmd", output_format = "html_document", output_dir = "/home/user")
```
The R Markdown gets rendered and stored in the output_dir. If you want to send it automatically to somebody, you need a tool or library that can be used from cron. I tested tools inside the linux world like Sendmail and postfix, but I was not a able to get it work. Same applied for the sendMailR Package. 

One important thing to do is to allow less secure apps get access to your mail account, which you can do here:
https://myaccount.google.com/lesssecureapps

I managed to get this code working with the following python script:

```python
#!/usr/bin/env python3

import smtplib
from email.mime.text import MIMEText
from email.mime.multipart import MIMEMultipart
from email.mime.base import MIMEBase
from email import encoders

email_user = 'youownmail@gmail.com'
email_password = 'password'
email_send = 'theothersmail.com'

subject = 'yourreport'

msg = MIMEMultipart()
msg['From'] = email_user
msg['To'] = email_send
msg['Subject'] = subject

body = 'Wuffi!'
msg.attach(MIMEText(body,'plain'))

filename='/home/user/stockreport.html'
attachment  =open(filename,'rb')

part = MIMEBase('application','octet-stream')
part.set_payload((attachment).read())
encoders.encode_base64(part)
part.add_header('Content-Disposition',"attachment; filename= "+ filename)

msg.attach(part)
text = msg.as_string()
server = smtplib.SMTP('smtp.gmail.com',587)
server.starttls()
server.login(email_user,email_password)

server.sendmail(email_user,email_send,text)
server.quit()

print("script done")
```

Run this script from cron and y

# Deploy Shiny apps on your own website
<!-- If you need PDF output, uncomment bookdown::pdf_book above in YAML. You will need a LaTeX installation, e.g., https://yihui.name/tinytex/ -->

# Deploy Shiny apps on your own website




<!-- If you need PDF output, uncomment bookdown::pdf_book above in YAML. You will need a LaTeX installation, e.g., https://yihui.name/tinytex/ -->
